---
title: "BIOST 534: Homework 1 "
author: "Allen Roberts"
date: "`r Sys.Date()`"
geometry: margin=1.75cm
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

1. As mentioned in the recorded lecture, using logarithms can help reduce numerical errors when working with large values that may not be easily represented by a computer. In this example, I calculated the log of the determinant of a matrix by summing the log of the eigenvalues of the matrix. This operation is expressed in the $\texttt{logdet}$ function shown below:

    ```{r warning = FALSE}
    logdet <- function(mat) {
    
      return(sum(log(eigen(mat, only.values = TRUE)$values)))
    
    }
  
    mat <- matrix(c(2, -1, 0, -1, 2, -1, 0, -1, 2), nrow = 3, ncol = 3)
  
    print(paste("My function's result:", logdet(mat)))
    print(paste("Result using R's log(det(mat)):", log(det(mat))))
    ```
    Using a simple 3x3 positive definite matrix, the $\texttt{logdet}$ function returns the same answer as taking the log of the built-in $\texttt{det}$ function in $\texttt{R}$. 
    To better understand why the $\texttt{logdet}$ function is a useful approach, I created a ``bad'' version of the function that takes the logarithm of the product of the eigenvalues of the matrix, as shown below:
    
    ```{r warning = FALSE}
    logdet_bad <- function(mat) {
      
      return(log(prod(eigen(mat, only.values = TRUE)$values)))
      
    }
    print(logdet_bad(mat))
    ```
    As the matrix dimension increases, the product of the eigenvalues will increase very rapidly, so this function may result in an overflow error for larger matrices. I created a large positive definite matrix using [this method](https://stats.stackexchange.com/questions/215497/how-to-create-an-arbitrary-covariance-matrix) as a test case to compare different methods for calculating the log determinant.
    ```{r warning = FALSE}
    set.seed(503)
    n <- 1000
    p <- qr.Q(qr(matrix(rnorm(n^2), n)))
    large_matrix <- crossprod(p, p*(n:1))
    
    print(paste("My function's result:", logdet(large_matrix)))
    print(paste("My bad function's result:", logdet_bad(large_matrix)))
    
    print(paste("Result using R's log(det(large_matrix)):",
                log(det(large_matrix))))
      print(paste("Result using R's determinant(large_matrix, logarithm = TRUE):",
                  determinant(large_matrix, logarithm = TRUE)$modulus[1]))
    ```
    As shown above, my $\texttt{logdet}$ function returns the correct determinant, as verified by using a built-in R function that also calculates the determinant on the log scale. In comparison, the version of my function that works with the product of the eigenvalues returns $\texttt{Inf}$, indicating that an overflow error has occurred. This also occurs if I take the log of the built-in $\texttt{det}$ function in R.

2. Using the equation given for the marginal likelihood $p(D_1, D_A | [1|A])$, the function below calculates the log marginal likelhood for any $A \subset \{2,...,p\}$. The function works on the additive scale using the logs of the terms in the right-hand side of the equation provided for $p(D_1, D_A | [1|A])$, thus avoiding potential numerical instability that might result when instead calculating the product of the terms. Note the function uses the $\texttt{logdet}$ function defined in problem 1 to calculate the log determinant. The log marginal likelihood calculation specified in the problem is calculated at the end.
    ```{r warning = FALSE}
    logmarglik <- function(data, A) {
      
      n <- nrow(data)
      
      D_1 <- as.matrix(data[, 1], nrow = n)
      D_A <- as.matrix(data[, A], nrow = n)
      M_A <- diag(length(A)) + t(D_A)%*%(D_A)
        
      lml <- lgamma((n + length(A) + 2)/2) - lgamma((length(A) + 2)/2) - 
        (1/2)*logdet(M_A) - ((n + length(A) + 2)/2) * 
        log((1 + t(D_1)%*%D_1 - t(D_1)%*%D_A%*%solve(M_A)%*%t(D_A)%*%D_1))
      
      return(as.numeric(lml))

    }
    
    erdata <- read.table("erdata.txt")
    print(paste("logmarglik(data, c(2,5,10)) = ", logmarglik(erdata, c(2,5,10))))
    ```

